Improving reproducibility of scientific software using Nix/NixOS: A case study on the preCICE ecosystem

### 2024-08-14 paper/paper.tex
feat: integrate last feedback

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/coreutils.png}
    \caption{The dependency graph of the coreutils package, generated by \texttt{nix-store --query --graph \$(nix build nixpkgs\#coreutils --print-out-paths) | dot -Tpng -Grankdir=LR -ocoreutils.png}}
    \label{fig:nix-graph}
\end{figure}

As Nix is purely functional, it relies on functions that, without side-effects, realize the package derivations.
Inputs of these functions are parameters such as the package version or the location of the source code.
Evaluating these functions with the same inputs yields the same outputs.
This is a critical factor for Nix's reproducibility.
Outputs, i.e. the build artifacts, never change after being built once.
Nix builds are, moreover, sandboxed, meaning that there is no internet access possible during a build.

All contents, including source files and resulting build artifacts, are stored inside the Nix store.
Per default, the Nix store resides in the path \texttt{/nix/store} on the file system.
The naming scheme for packages includes the above mentioned hash, for preCICE v2.5.0, for example,
\path{/nix/store/0a5gw3l...-precice-2.5.0}.
This eases checking the Nix store for integrity and ensures that builds using the same inputs are performed only once.
As all build outputs reference their complete dependency graphs inside the Nix store, they do not interfere with other build outputs, enabling installation of multiple versions of the same software without conflicts.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/precice.png}
    \caption{Software setup of preCICE from the preCICE website \url{https://precice.org}}
    \label{fig:precice}
\end{figure}

@@

We present challenges, solutions, and workarounds for packaging all components of the preCICE ecosystem.
The Nix code for all components is available online\footnote{\url{https://github.com/precice/nix-packages/releases/tag/deRSE24-paper-submission}}.

@@

The \texttt{pyprecice} package is already available upstream, but does not compile as it is not able to find the python module \texttt{pkgconfig}.
After adding this single dependency as an input to the package definition, the python package builds and can be used as expected.
This solution was contributed to the upstream nixpkgs repository\footnote{\url{https://github.com/NixOS/nixpkgs/commit/fd8962162ac21be59fc3a05fb6a250eeab6b2bec}}.

@@

The remaining two dependencies, Mumps and Tfel are not available upstream and need to be packaged.
Mumps uses a Makefile and can be configured using Makefile.inc.
We write our own custom Makefile.inc by providing the locations for dependencies, such as Scotch, METIS, ParMETIS, or Blacs.
Almost all dependencies are available upstream, but we need to recompile Scotch with additional build flags.
Moreover, for packaging Blacs, which also uses a Makefile, we configure the build using a custom \texttt{Bmake.inc} file, which is used to set compiler flags, the path to MPI, and the bash installation.
Tfel, on the other hand, uses CMake as build system making it easy to package it with Nix.

@@

\begin{figure*}
    \normalsize
    \begin{minted}{nix}
{ lib, stdenv, fetchFromGitHub, cmake, precice,
  dealii, enable3d ? false }:
stdenv.mkDerivation rec {
  pname = "precice-dealii-adapter";
  version = "unstable-2022-09-23"; # could also be a git tag

@@

  nativeBuildInputs = [ cmake ]; # build-time dependencies
  buildInputs = [ precice dealii ]; # run-time dependencies

@@

  # nix' default phases can be overwritten such as:
  installPhase = ''
    mkdir -p $out/bin && cp elasticity $out/bin/elasticity
  '';
}
    \end{minted}
    \caption{Nix code to package the deal.II-preCICE adapter}
    \label{lst:dealii-adapter-nix}
\end{figure*}

@@

FEniCS~\cite{fenics} is already packaged upstream, however, not with all features the FEniCs-preCICE adapter~\cite{Rodenberg2021} relies on, in particular PETSc support and mshr, the FEniCS mesh generator.
This is why we need to package PETSc4py, the python bindings package of PETSc, which is not available upstream.
The build process uses the internal \texttt{buildPythonApplication} build tool and runs successfully once we add the option \texttt{build\_src --force} to rebuild cython code and pin the cython version to \texttt{0.29.34}~\footnote{\url{https://gitlab.com/petsc/petsc/-/issues/1359}}.
We could not enable tests for PETSc4py, however, because they depend on OpenMPI and the network, which is not available within Nix's build sandbox.
Afterwards, we specify PETSc4py as dependency for FEniCS and enable the PETSc support.
Additionally, we need to recompile PETSc with additional features enabled that are not enabled by the upstream Nix package. These include ParMETIS, HYPRE and ScaLAPACK. We use \texttt{overrideAttrs}, a Nix feature that allows us to change an existing package.
Lastly, we also package mshr, so all required features for the adapter package are now are available.

@@

OpenFOAM uses its own custom build system called \texttt{wmake}, which is typically called with a \texttt{Allwmake} wrapper script.
The build system sets 36 environment variables, one step at a time by checking several parameters, e.g.~the CPU architecture or the location of the source code. Also during runtime, these variables need to be set, either by a wrapper script or by manually sourcing a file inside the installation directory of OpenFOAM.
For Nix, these properties are unfortunate. To compile OpenFOAM with Nix, we need to patch the shebangs\footnote{\url{https://foldoc.org/shebang}} of wmake to make it run during the build.
We also use a shell script which exports the environment variables to the current shell session.
For simplicity, we hard-code all parameters, such as the processor architecture. These could be parametrized, however, based on the Nix inputs to allow for optimized builds. We use the script during the build phase to source all variables such that, for example, \texttt{OPENFOAM\_SRC\_PATH} points to the default location \texttt{/build/openfoam}.
Afterwards, \texttt{./Allwmake -j -q} is sufficient to start the build.
The install phase then copies the necessary files and directories to \texttt{\$out}, replaces the mock \texttt{OPENFOAM\_SRC\_PATH} by the value of \texttt{\$out}, and creates a wrapper for the \texttt{openfoam} shell script.

@@

The current preCICE VM\footnote{\url{https://precice.org/installation-vm.html}} is built on top of Vagrant\footnote{\url{https://www.vagrantup.com/}} using Ubuntu as a base image.
When provisioned for the first time, it further installs software, compiles programs, and clones the preCICE tutorials\footnote{\url{https://github.com/precice/tutorials/}}.
This means that we only have reproducibility when fetching the VM from vagrant cloud~\footnote{\url{https://app.vagrantup.com/precice/boxes/precice-vm}}, as we can not rebuild the VM from scratch, because the provision step currently fetches files from the internet that are at this very point in time the \textit{latest} version, i.e. the \texttt{main} branch of the preCICE repository rather than a pinned tag or commit.
Making us dependent on vagrant cloud to continue to serve the box.

@@

Finally, hard-coding libraries to \texttt{/usr/lib} and \texttt{/usr/include} might work on one system, but might make it hard or even impossible to install a package on a different system.
Configurable solutions, such as \texttt{pkg-config}, should always be preferred and work out of the box with common paths such as \texttt{/usr/lib}.

@@

Out of the 20 components of the preCICE distribution, we were able to package 14 components ourselfs, four were already packaged upstream and two were not packageable as described earlier.
Additionally we were able to run 52 participant examples out of the 61 total examples found in the preCICE tutorials, showing the potential of Nix in the domain of scientific software.
Many packages required workarounds, however, which might be difficult to achieve for non-experienced Nix users.
Nix also comes with a few peculiarities, which complicate workarounds.
In fact, every piece of software has to lie within \texttt{/nix/store} and each path therein is its own-isolated tree inspired by the Filesystem Hierarchy Standard (FHS).

>>>>

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/coreutils.png}
    \caption{The full dependency graph of the coreutils package, generated by \texttt{nix-store --query --graph \$(nix build nixpkgs\#coreutils --print-out-paths) | dot -Tpng -Grankdir=LR -ocoreutils.png}}
    \label{fig:nix-graph}
\end{figure}

As Nix is purely functional, it relies on functions that, without side-effects, realize the package derivations.
Inputs of these functions are parameters such as the package version, the location of the source code and its dependencies.
Nix then offers built-in functions for common buildsystems and languages such as cmake, rust, go and python by implementing different phases.
There are a wide range of phases that always get executed in the same order and package maintainers can then decide weather the default phase is sufficient or needs to be overriden in order to produce a working build.
The most common phases to override are the \texttt{configurePhase}, which is mostly used for cmake or autotools configuration, the \texttt{buildPhase}, \texttt{installPhase} and the \texttt{checkPhase} which can for example be used to run unit tests.
It is also worth mentioning that most phases have a pre and post step, which users can use to do slight adjustments when they need to override excisting packages.
Evaluating these functions with the same inputs yields the same outputs.
This is a critical factor for Nix's reproducibility.
Outputs, i.e. the build artifacts, never change after being built once.
Nix builds are, moreover, sandboxed, meaning that there is no internet access possible during a build.

All contents, including source files and resulting build artifacts, are stored inside the Nix store.
Per default, the Nix store resides in the path \path{/nix/store} on the file system.
The naming scheme for packages includes the above mentioned hash, for preCICE v2.5.0, for example,
\path{/nix/store/0a5gw3l...-precice-2.5.0}.
This eases checking the Nix store for integrity and ensures that builds using the same inputs are performed only once.
As all build outputs reference their complete dependency graphs inside the Nix store, they do not interfere with other build outputs, enabling installation of multiple versions of the same software without conflicts.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/precice.png}
    \caption{Software setup of preCICE from the preCICE website \url{https://precice.org}. It shows how a custom solver can be coupled using an adapter with existing adapters for OpenFOAM and CalculiX.}
    \label{fig:precice}
\end{figure}

@@

In this section we will look and discuss all components within the preCICE ecosystem.
We then present challenges, solutions and workarounds that occur when packaging these components.
This section is structured as follows, we first look into preCICE itself, followed by additional tooling and bindings of the preCICE library in other languages.
Afterwards we present solutions for all solvers and their respected adapters.
Lastly we show that Nix additionally be utilized to generate iso files as well as qemu and Vagrant VM images.
The Nix code for all components is available online\footnote{\url{https://github.com/precice/nix-packages/releases/tag/deRSE24-paper-submission}} and can be used to download and run any of the presented packages.

@@

The pyprecice package is already available upstream, but does not compile as it is not able to find the python module pkgconfig.
After adding this single dependency as an input to the package definition, the python package builds and can be used as expected.
This solution was contributed to the upstream nixpkgs repository\footnote{\url{https://github.com/NixOS/nixpkgs/commit/fd8962162ac21be59fc3a05fb6a250eeab6b2bec}}.

@@

The remaining two dependencies, Mumps and Tfel are not available upstream and need to be packaged.
Mumps uses a Makefile and can be configured using \texttt{Makefile.inc}.
We write our own custom Makefile.inc by providing the locations for dependencies, such as Scotch, METIS, ParMETIS, or Blacs.
Almost all dependencies are available upstream, but we need to recompile Scotch with additional build flags.
Moreover, for packaging Blacs, which also uses a Makefile, we configure the build using a custom \texttt{Bmake.inc} file, which is used to set compiler flags, the path to MPI, and the bash installation.
Tfel, on the other hand, uses CMake as build system making it easy to package it with Nix.

@@

\begin{figure*}
    \normalsize
    \begin{minted}{nix}
{
  # Inputs including nix functions/packages or additional flags
  lib, stdenv, fetchFromGitHub, cmake,  dealii,
  # Optional features that users can enable when overriding
  enable3d ? false }:
stdenv.mkDerivation rec {
  pname = "precice-dealii-adapter";
  version = "unstable-2022-09-23"; # could also be a git tag

@@

  # dependencies only available at build-time
  nativeBuildInputs = [ cmake ];
  # dependencies that also need to be installed at run-time
  # mostly libraries needed for linking and for execution
  buildInputs = [ precice dealii ];

@@

  # nix' default phases can be overwritten such as:
  installPhase = ''
    # $out contains the final path inside the nix store.
    # The resulting build files are copied into this directory
    mkdir -p $out/bin && cp elasticity $out/bin/elasticity
  '';
}
    \end{minted}
    \caption{Nix code to package the deal.II-preCICE adapter}
    \label{lst:dealii-adapter-nix}
\end{figure*}

@@

FEniCS~\cite{fenics} is already packaged upstream, however, not with all features the FEniCs-preCICE adapter~\cite{Rodenberg2021} relies on, in particular PETSc support and mshr, the FEniCS mesh generator.
This is why we need to package PETSc4py, the python bindings package of PETSc, which is not available upstream.
The build process uses the internal \texttt{buildPythonApplication} build tool and runs successfully once we add the option \texttt{build\_src --force} to rebuild cython code and pin the cython version to 0.29.34~\footnote{\url{https://gitlab.com/petsc/petsc/-/issues/1359}}.
We could not enable tests for PETSc4py, however, because they depend on OpenMPI and the network, which is not available within Nix's build sandbox.
Afterwards, we specify PETSc4py as dependency for FEniCS and enable the PETSc support.
Additionally, we need to recompile PETSc with additional features enabled that are not enabled by the upstream Nix package. These include ParMETIS, HYPRE and ScaLAPACK. We use \texttt{overrideAttrs}, a Nix feature that allows us to change an existing package.
Lastly, we also package mshr, so all required features for the adapter package are now are available.

@@

OpenFOAM uses its own custom build system called wmake, which is typically called with a Allwmake wrapper script.
The build system sets 36 environment variables, one step at a time by checking several parameters, e.g.~the CPU architecture or the location of the source code. Also during runtime, these variables need to be set, either by a wrapper script or by manually sourcing a file inside the installation directory of OpenFOAM.
For Nix, these properties are unfortunate. To compile OpenFOAM with Nix, we need to patch the shebangs\footnote{\url{https://foldoc.org/shebang}} of wmake to make it run during the build.
We also use a shell script which exports the environment variables to the current shell session.
For simplicity, we hard-code all parameters, such as the processor architecture. These could be parametrized, however, based on the Nix inputs to allow for optimized builds. We use the script during the build phase to source all variables such that, for example, \texttt{OPENFOAM\_SRC\_PATH} points to the default location \texttt{/build/openfoam}.
Afterwards, \texttt{./Allwmake -j -q} is sufficient to start the build.
The install phase then copies the necessary files and directories to \texttt{\$out}, replaces the mock \texttt{OPENFOAM\_SRC\_PATH} by the value of \texttt{\$out}, and creates a wrapper for the \texttt{openfoam} shell script.

@@

The current preCICE VM\footnote{\url{https://precice.org/installation-vm.html}} is built on top of Vagrant\footnote{\url{https://www.vagrantup.com/}} using Ubuntu as a base image.
When provisioned for the first time, it further installs software, compiles programs, and clones the preCICE tutorials\footnote{\url{https://github.com/precice/tutorials/}}.
This means that we only have reproducibility when fetching the VM from vagrant cloud~\footnote{\url{https://app.vagrantup.com/precice/boxes/precice-vm}}, as we can not rebuild the VM from scratch, because the provision step currently fetches files from the internet that are at this very point in time the \textit{latest} version, i.e. the main branch of the preCICE repository rather than a pinned tag or commit.
Making us dependent on vagrant cloud to continue to serve the box.

@@

Finally, hard-coding libraries to \path{/usr/lib} and \path{/usr/include} might work on one system, but might make it hard or even impossible to install a package on a different system.
Configurable solutions, such as pkg-config, should always be preferred and work out of the box with common paths such as \path{/usr/lib}.

@@

Out of the 20 components of the preCICE distribution, we were able to package 14 components ourselfs, four were already packaged upstream and two were not packageable as described earlier.
Additionally we were able to run 52 participant examples out of the 61 total examples found in the preCICE tutorials, showing the potential of Nix in the domain of scientific software.
Many packages required workarounds, however, which might be difficult to achieve for non-experienced Nix users.
Nix also comes with a few peculiarities, which complicate workarounds.
In fact, every piece of software has to lie within \path{/nix/store} and each path therein is its own-isolated tree inspired by the Filesystem Hierarchy Standard (FHS).
